Cognate beginnings to bilingual lexical acquisition

### 2023-07-25 manuscript/manuscript.qmd
Update Dockerfile

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts. Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @vanhell2008sentence; @hoshino2008cognate; @morford2011when; @schwartz2007reading; @spivey1999cross; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical].

@@

>>>>

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts. Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @vanhell2008sentence; @hoshino2008cognate; @morford2011when; @schwartz2007reading; @spivey1999cross; @shook2012bimodal; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @floccia2020translation; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical].

@@

The asymmetric facilitation of cognateness on word acquisition reported in the present study parallels previous findings in toddlers and adults. For instance, unbalanced (or low-proficiency) bilinguals benefit from forward priming (dominant to non-dominant) cross-language phonological priming during word processing [for toddlers see @vonholzen2019impact; @singh2014one; for adults see @grainger1998masked; @voga2007cognate; @de1991lexical; @shook2019covert; @dimitropoulou2011masked; but see @jardak2019labels]. One the other hand, backward priming (non-dominant to dominant) seems less robust and a more challenging effect to detect [e.g., @finkbeiner2004role; @hoshino2010erp; @midgley2009language; but see @duyck2009translation; @schoonbaert2009semantic]. More balanced (or high-proficiency bilinguals) show an equivalent priming facilitation in both directions [@basnight2007differences; @dunabeitia2009masked]. These results have been taken as evidence for an asymmetry in the strength of forward and backward connections in the unbalanced bilingual lexicon.

Although implemented in different ways, or found under different assumptions, a dominance-mediated asymmetry is accounted for by previous models of adult bilingual lexical access, like the Revised Hierarchical Model [@kroll1994category], BIA/BIA+ [@dijkstra2013bia; @dijkstra2002architecture], BLINCS [@shook2013bilingual], or Multilink [@dijkstra2019multilink], and also by models providing a more development-oriented perspective into how lexical representations in a second language are formed, like the Ontogenic Model [@bordag2022ontogenesis; @cook2016fuzzy], and BIA-d [@grainger2010re]. This seems to provide a convenient account for the interaction between language dominance and cognateness found in the present study. Following a similar pattern to those predicted by the aforementioned models, infants benefited from the cognate status of only words in their low-dominance language, and especially if their exposure to both languages was unbalanced.

These models, however, are aimed at explaining results in adults, and their predictions should be taken with a grain of salt when extended to early language acquisition. Language dominance and proficiency are frequently defined in both populations using dimensions other than degree of exposure, which is the most common practice in infant research [@marian2021measuring; @rocha2023defining]. For instance, the low-proficiency bilinguals in many of the aforementioned studies acquired their second language years after their toddlerhood. This has three important consequences for their assumed developmental trajectories of language acquisition. First, the acquisition of the phonological system of the new language must be negotiated with the already acquired phoneme inventory of the first language [e.g., @cutler2006asymmetric; @sebastian2005influence; @sebastian2006first], which is in place around the first year of life [see @werker2015critical for review]. Second, adults acquiring a second language already possess a system of word-meaning mappings, while simulatenous bilingual infants must generate a lexicon in two languages in the absence of clear word-meaning mappings. Third, subjects are assumed to be literate and possess an orthographic system in place, which may shape how new words are integrated in the lexicon and processed during experimental tasks [e.g., @thierry2007brain]. In this scenario, the acquisition of a second language may take place in a substantially different way compared to how bilingual infants acquire two languages from birth.

The DevLex-II model [@zhao2010bilingual] captures some of these features of the early bilingual lexicon, and considers the case of bilinguals in the process of stimultaneously acquiring their two languages. In line with the aforementioned models, DevLex-II predicts asymmetries between the dominant and the non-dominant language that arise from a difference in the frequency of the input, or the order of acquisition of the non-dominant language (even delayed by only some months). For instance, simulations from DevLex-II result in an asymmetric cross-language priming, in which words from the dominant (acquired acquired) language primed more strongly the recognition of words in the non-dominant language (later acquired) than in the other direction [@zhao2013simulating]. In addition, the simulated lexica showed sparser connectivity between words in the non-dominant language, compared to the connectivity between words in the dominant language.

 As suggested in the introduction, it is possible that the asymmetric effect of cognateness found in the present study is simply the result of infants being exposed more frequently to words in the dominant language. This would lead to words in the non-dominant language receiving additional parallel activation, compared to words in the dominant language, and therefore benefiting more strongly from their cognate status. This explanation does not necessarily compete with words in the dominant language activating more *strongly* their translations than *vice versa*. Further research is needed in order to clarify this issue.
